---
title: "Materiality Heat Map"
author: "Sam Hickman"
format: html
editor: visual
---

If organisations are looking to adapt this workflow, go to all of the `# CHANGE HERE` flags and ensure these are changed to suit your organisations specific needs

# Setup workflow

## Load packages

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(janitor)
  library(readxl)
  library(writexl)
  library(patchwork)
  library(plotly)
})
```

## Load functions, colours, and factors

```{r}
source("setup.R")
```

## Load data

```{r}
# Import procurement data
procurement <- read_csv("procurement/dummy_procurement_data.csv") # CHANGE HERE

# Crosswalk table that maps the Australasian Universities Procurement Network (AUPN)
# economic classifications to The UN's ISIC Rev.4 econonic classifications.
# If an organisation does not use AUPN, or an adapted version of AUPN,
# this can be changed to meet that organisations specific classification.
procurement_to_isic <- read_excel("procurement/aupn_to_isic.xlsx")

# Imports ENCORE pressure data
encore_pressure <- read_csv("encore_download/clean/encore_pressure_ratings.csv")

# Imports ENCORE dependency data
encore_dependency <- read_csv("encore_download/clean/encore_dependency_ratings.csv")

# Imports upstream value chain links
upstream_links <- read_csv("encore_download/clean/upstream_links.csv")

# Imports downstream value chain links
downstream_links <- read_csv("encore_download/clean/downstream_links.csv")
```

# Add ISIC classification and materiality ratings to procurement data

## Add the ISIC classifications to the procurement data

```{r}
# Create a column called Mega_L1_2_3 that can be used to map the procurement spend
# to the ISIC Classifications. The reason Mega_L1_2_3 is created is that there
# are duplicate L3 titles, so we must combine Mega, L1, L2, and L3 to ensure accurate matching.
# This works for data that is in the AUPN classification

# 1. Create a combined key within the procurement data
procurement_mapping <- procurement |>
  mutate(
    Mega_L1_2_3 = str_c(
      `Mega Category (Taxonomy)`,
      `Level 1 (Taxonomy)`,
      `Level 2 (Taxonomy)`,
      `Level 3 (Taxonomy)`,
      sep = "|"
    )
  ) |>
  select(
    `Supplier Name`,
    `Mega Category (Taxonomy)`,
    `Level 1 (Taxonomy)`,
    `Level 2 (Taxonomy)`,
    `Level 3 (Taxonomy)`,
    Spend,
    Mega_L1_2_3
  )

# 2. Create the same combined key in the AUPN to ISIC crosswalk table
isic_mapping <- procurement_to_isic |>
  mutate(
    Mega_L1_2_3 = str_c(
      `Mega Category (Taxonomy)`,
      `Level 1 (Taxonomy)`,
      `Level 2 (Taxonomy)`,
      `Level 3 (Taxonomy)`,
      sep = "|"
    )
  ) |>
  select(
    Mega_L1_2_3,
    `ISIC Unique Code`,
    `ISIC Section`,
    `ISIC Division`,
    `ISIC Group`,
    `ISIC Class`,
    `Match Quality`,
    Comments
  )

# 3. Join the procurement data with the ISIC crosswalk table
procurement_to_isic <- left_join(
  procurement_mapping,
  isic_mapping,
  by = "Mega_L1_2_3"
) |>
  select(-Mega_L1_2_3) # remove the merged key if you don't need it anymore

# Check for duplicate rows? Zero using the dummy data
nrow(procurement_to_isic) - nrow(procurement_mapping)
```

## Add the materiality ratings to the procurement data

```{r}
# Adding pressure ratings to procurement
procurement_pressure <- procurement_to_isic |>
  inner_join(
    encore_pressure,
    by = "ISIC Unique Code",
    suffix = c("", ".y")
  ) |>
  select(
    -`ISIC Section.y`,
    -`ISIC Division.y`,
    -`ISIC Group.y`,
    -`ISIC Class.y`,
    -`Match Quality`,
    -`Comments`
  )

# Duplicate rows?
nrow(procurement_pressure) - nrow(procurement_to_isic)

# Adding dependency ratings to procurement
procurement_dependency <- procurement_to_isic |>
  inner_join(
    encore_dependency,
    by = "ISIC Unique Code",
    suffix = c("", ".y")
  ) |>
  select(
    -`ISIC Section.y`,
    -`ISIC Division.y`,
    -`ISIC Group.y`,
    -`ISIC Class.y`,
    -`Match Quality`,
    -`Comments`
  )

# Duplicate rows?
nrow(procurement_pressure) - nrow(procurement_to_isic)
```

# Create heatmaps

## Pressure heatmap

### Additional instructions

To exclude certain ENCORE “pressures” that are immaterial or reduce the clarity of the visual, add them to the `!Category` section of the script.

To filter out ISIC Sections with very small expenditures, uncomment the line
`Spend > 2e6, # CHANGE HERE`
and adjust the threshold to a value that is appropriate for your organisation.

```{r}
# 1. Aggregate worst-case pressure scores and total spend per ISIC Section
summarised_pressure_data <- procurement_pressure |>
  group_by(`ISIC Section`) |>
  summarise(
    across(where(is.numeric), ~ sum(.x, na.rm = TRUE)),
    across(all_of(pressure_score_cols), ~ pick_highest_level(.x, score_levels)),
    .groups = "drop"
  ) |>
  select(`ISIC Section`, Spend, all_of(pressure_score_cols))

# 2. Reshape to long format and compute share of total spend
summarised_pressure_data_long <- summarised_pressure_data |>
  pivot_longer(
    cols = -(c(Spend, `ISIC Section`)),
    names_to = "Category",
    values_to = "Value"
  ) |>
  mutate(
    percentage_spend = Spend / sum(summarised_pressure_data$Spend, na.rm = TRUE) * 100
  )

# 3. Apply materiality threshold and drop irrelevant categories
pressure_plot_dat <- summarised_pressure_data_long |>
  filter(
    # Spend > 2e6, # CHANGE HERE -> Alter threshold for heatmap to only show the most important categories
    !Category %in%
      c(
        # CHANGE HERE
        # Remove pressure categories that are less relevant to your organisation
        # e.g. "Other abiotic resource extraction"
      )
  ) |>
  mutate(
    percentage_spend = if_else(duplicated(percentage_spend), 0, percentage_spend),
    Value = if_else(Value %in% c("N/A", "ND"), "ND", Value)
  ) |>
  arrange(desc(Spend)) |>
  mutate(
    `ISIC Section` = factor(`ISIC Section`, levels = rev(unique(`ISIC Section`))),
    Category = factor(Category, levels = pressure_score_cols)
  )

# 4. Heat-map and spend bar
p_heatmap <- ggplot(pressure_plot_dat, aes(Category, `ISIC Section`, fill = Value)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  scale_fill_manual(
    values = pressure_colors,
    drop = FALSE,
    breaks = c("ND", "VL", "L", "M", "H", "VH"),
    labels = c("No Data", "Very Low", "Low", "Medium", "High", "Very High"),
    name = "Materiality\nScore"
  ) +
  labs(x = "Pressure Category", y = "Spend Category (ISIC Rev.4 Section)") +
  scale_x_discrete(labels = \(x) str_wrap(x, 20)) +
  scale_y_discrete(labels = \(x) str_wrap(x, 25)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  guides(fill = guide_legend(title.position = "left", nrow = 1, byrow = TRUE))

p_bar <- ggplot(pressure_plot_dat, aes(percentage_spend, `ISIC Section`)) +
  geom_col(fill = "#C8C8C8") +
  labs(x = "Percentage of Total\nProcurement Spend", y = NULL) +
  scale_y_discrete(labels = \(x) str_wrap(x, 20)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank()
  )

combined_plot_p <- p_heatmap + p_bar + plot_layout(widths = c(5, 1))

print(combined_plot_p)

ggsave(
  "sample_outputs/sample_pressure_plot.pdf",
  combined_plot_p,
  width = 15,
  height = 10,
  dpi = 300
)
```

## Dependency heatmap

### Additional instructions

To exclude certain ENCORE “ecosystem services” that are immaterial or reduce the clarity of the visual, add them to the `!Category` section of the script.

To filter out ISIC Sections with very small expenditures, uncomment the line
`Spend > 2e6, # CHANGE HERE`
and adjust the threshold to a value that is appropriate for your organisation.

```{r}
# 1. Aggregate worst-case dependency scores and total spend per ISIC Section
summarised_dependency_data <- procurement_dependency |>
  group_by(`ISIC Section`) |>
  summarise(
    across(where(is.numeric), ~ sum(.x, na.rm = TRUE)),
    across(all_of(dependency_score_cols), ~ pick_highest_level(.x, score_levels)),
    .groups = "drop"
  ) |>
  select(
    `ISIC Section`,
    Spend,
    all_of(dependency_score_cols)
  )

# 2. Reshape to long format and compute share of total spend
summarised_dependency_data_long <- summarised_dependency_data |>
  pivot_longer(
    cols = -(c(Spend, `ISIC Section`)),
    names_to = "Category",
    values_to = "Value"
  ) |>
  mutate(
    percentage_spend = Spend / sum(summarised_dependency_data$Spend, na.rm = TRUE) * 100
  )

# 3. Apply materiality threshold and drop irrelevant categories
dependency_plot_dat <- summarised_dependency_data_long |>
  filter(
    # Spend > 2e6, # CHANGE HERE -> Alter threshold for heatmap to only show the most important categories
    !Category %in%
      c(
        # CHANGE HERE
        # Remove dependency categories that are less relevant to your organisation
        "Nursery population and habitat maintenance",
        "Other provisioning services - Animal-based energy",
        "Other regulating and maintenance service - Dilution by atmosphere and ecosystems",
        "Other regulating and maintenance service - Mediation of sensory impacts (other than noise)"
      )
  ) |>
  mutate(
    percentage_spend = if_else(duplicated(percentage_spend), 0, percentage_spend),
    Value = if_else(Value %in% c("N/A", "ND"), "ND", Value)
  ) |>
  arrange(desc(Spend)) |>
  mutate(
    `ISIC Section` = factor(`ISIC Section`, levels = rev(unique(`ISIC Section`))),
    Category = factor(Category, levels = dependency_score_cols)
  )

# 4. Heat-map and spend bar
d_heatmap <- ggplot(dependency_plot_dat, aes(Category, `ISIC Section`, fill = Value)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  scale_fill_manual(
    values = pressure_colors,
    drop = FALSE,
    breaks = c("ND", "VL", "L", "M", "H", "VH"),
    labels = c("No Data", "Very Low", "Low", "Medium", "High", "Very High"),
    name = "Materiality\nScore"
  ) +
  labs(x = "Ecosystem Service", y = "Spend Category (ISIC Rev.4 Section)") +
  scale_x_discrete(labels = \(x) str_wrap(x, 20)) +
  scale_y_discrete(labels = \(x) str_wrap(x, 25)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal"
  ) +
  guides(fill = guide_legend(title.position = "left", nrow = 1, byrow = TRUE))

d_bar <- ggplot(dependency_plot_dat, aes(percentage_spend, `ISIC Section`)) +
  geom_col(fill = "#C8C8C8") +
  labs(x = "Percentage of Total\nProcurement Spend", y = NULL) +
  scale_y_discrete(labels = \(x) str_wrap(x, 20)) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank()
  )

combined_plot_d <- d_heatmap + d_bar + plot_layout(widths = c(6, 1))
print(combined_plot_d)

ggsave(
  "sample_outputs/sample_dependency_plot.pdf",
  combined_plot_d,
  width = 20,
  height = 10,
  dpi = 300
)
```

# Calculate materialities by supplier

```{r}
# Process procurement_pressure dataset with worst-case materiality aggregation
procurement_pressure_with_supplier_total <- procurement_pressure |>
  group_by(
    `Supplier Name`,
    `Level 1 (Taxonomy)`,
    `Level 2 (Taxonomy)`,
    `Level 3 (Taxonomy)`,
    `ISIC Unique Code`
  ) |>
  summarise(
    `Total Spend` = sum(Spend, na.rm = TRUE),
    across(all_of(pressure_score_cols), ~ pick_highest_level(.x, score_levels)),
    .groups = "drop"
  ) |>
  select(
    `Supplier Name`,
    `Total Spend`,
    everything()
  ) |>
  arrange(desc(`Total Spend`))

# Process procurement_dependency dataset with worst-case materiality aggregation
procurement_dependency_with_supplier_total <- procurement_dependency |>
  group_by(
    `Supplier Name`,
    `Level 1 (Taxonomy)`,
    `Level 2 (Taxonomy)`,
    `Level 3 (Taxonomy)`,
    `ISIC Unique Code`
  ) |>
  summarise(
    `Total Spend` = sum(Spend, na.rm = TRUE),
    across(all_of(dependency_score_cols), ~ pick_highest_level(.x, score_levels)),
    .groups = "drop"
  ) |>
  select(
    `Supplier Name`,
    `Total Spend`,
    everything()
  ) |>
  arrange(desc(`Total Spend`))
```

# Write all data into an Excel file

```{r}
# Write excel file of all data
write_xlsx(
  list(
    "Procurement data" = procurement,
    "Encore pressure ratings" = encore_pressure,
    "Encore dependency ratings" = encore_dependency,
    "Upstream links" = upstream_links,
    "Downstream links" = downstream_links,
    "AUPN-ISIC crosswalk" = procurement_to_isic,
    "Procurement & pressure" = procurement_pressure,
    "Procurement & dependency" = procurement_dependency,
    "Supplier & pressure" = procurement_pressure_with_supplier_total,
    "Supplier & dependency" = procurement_dependency_with_supplier_total
  ),
  path = "sample_outputs/Sample Materiality Workbook.xlsx"
)
```